CF Registry Analysis - Lung Function

### 2025-12-23 Christine Chan/CF Registry/cf_registry_lung_function_analysis.qmd
a bunch of updates for Christine

# Calculate number of visits (just for table 1)
n_vis = annual |>
  summarise(years_with_data = n())
cfrd_only <- left_join(cfrd_only, n_vis, by = join_by(eDWID))

@@

```{r}
# Create a model dataframe
mod_df = inner_join(
  annual,
  cfrd_only |> filter(dx_year >= 2004),
  by = join_by(eDWID)
) |>
  drop_na(a1c_ogtt) |>
  mutate(years_to_cfrd = ReviewYear - first_cfrd) |>
  filter(years_to_cfrd <= 5 & years_to_cfrd >= -5) |>
  mutate(years_to_cfrd = factor(years_to_cfrd))
var_label(mod_df$years_to_cfrd) = "Years From CFRD Dx"
# Write data for Christine to look at
mod_df |>
  select(
    eDWID,
    Race,
    Hispanicrace,
    MutClass,
    ReviewYear,
    Age_YrEnd,
    CFRD,
    first_cfrd,
    age_dx,
    A_weight,
    A_height,
    bmi_adult,
    bmi_perc,
    bmi_cat,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    a1c_ogtt,
    A_hgba1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    A_IsOnEnzymes,
    liver_disease,
    A_pseudomonasaeruginosa,
    A_fungalyeast1,
    A_mycocultureresults3,
    A_cultureresults3,
    treatment,
    cfrd_treatment_options2,
    cfrd_treatment_options3,
    cfrd_treatment_options4,
    years_with_data
  ) %>%
  write.csv(
    .,
    file = "./Data_Clean/fev1_model_data_2004.csv",
    row.names = FALSE,
    na = ""
  )
# Tables
mod_df |>
  filter(ReviewYear == first_cfrd) |>
  select(
    a1c_ogtt,
    age_dx,
    A_weight,
    A_height,
    bmi_adult,
    bmi_perc,
    bmi_cat,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    Race,
    Hispanicrace,
    MutClass,
    a1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    A_IsOnEnzymes,
    liver_disease,
    A_pseudomonasaeruginosa,
    A_fungalyeast1,
    A_mycocultureresults3,
    A_cultureresults3,
    treatment,
    cfrd_treatment_options2,
    cfrd_treatment_options3,
    cfrd_treatment_options4,
    years_with_data
  ) |>
  tbl_summary(
    by = a1c_ogtt,
    missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) |>
  add_overall() |>
  add_p(
    test = list(
      all_continuous() ~ "oneway.test",
      "years_with_data" ~ "kruskal.test"
    )
  ) |>
  modify_caption("**All Groups**")
mod_df |>
  filter(ReviewYear == first_cfrd, a1c_ogtt %in% c("OGTT", "HbA1c Only")) |>
  mutate(a1c_ogtt = droplevels(a1c_ogtt)) |>
  select(
    a1c_ogtt,
    age_dx,
    A_weight,
    A_height,
    A_bmivalue,
    A_bmipercentile,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    Race,
    Hispanicrace,
    MutClass,
    a1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    A_IsOnEnzymes,
    liver_disease,
    A_pseudomonasaeruginosa,
    A_fungalyeast1,
    A_mycocultureresults3,
    A_cultureresults3,
    treatment,
    cfrd_treatment_options2,
    cfrd_treatment_options3,
    cfrd_treatment_options4,
    years_with_data
  ) |>
  tbl_summary(
    by = a1c_ogtt,
    missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) |>
  add_overall() |>
  add_p(test = list(all_continuous() ~ "oneway.test")) |>
  modify_caption("**OGTT vs. HbA1c Only**")
```

@@

### LOESS-smoothed plot

```{r}
#| warning: false
#| message: false
# Plot by diagnosis type
ggplot(
  mod_df,
  aes(
    y = A_FEV1_pct_predicted,
    x = years_to_cfrd,
    color = a1c_ogtt,
    group = a1c_ogtt
  )
) +
  xlab("Years from CFRD") +
  geom_smooth() +
  theme_bw()
```

### Linear model results

#### Not adjusted for treatment status

Adjusted for age at diagnosis, sex, mutation class, and pancreatic enzyme use.

```{r}
# Fit the model
m <- lmer(
  A_FEV1_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    age_dx +
    Gender +
    MutClass +
    A_IsOnEnzymes +
    (1 | eDWID),
  data = mod_df
)
t3 = car::Anova(m, type = 'III', contrasts = "contr.sum")
emm = emmeans(m, specs = ~ years_to_cfrd + a1c_ogtt, pbkrtest.limit = 1e5)
emm = data.frame(emm)
ggplot(
  emm,
  aes(x = years_to_cfrd, y = emmean, group = a1c_ogtt, color = a1c_ogtt)
) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.25) +
  geom_line() +
  xlab("Years From CFRD Dx") +
  ylab("Estimated Mean ppFEV1") +
  labs(color = "Diagnosis Method") +
  theme_bw()
# Model tables
kable(emm, digits = 3)
m |> tbl_regression() |> add_n(location = "level")
t3
```

@@

#### Adjusted for treatment status

Adjusted for age at diagnosis, sex, mutation class, pancreatic enzyme use, and CFRD treatment.

```{r}
# Fit the model
m <- lmer(
  A_FEV1_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    age_dx +
    Gender +
    MutClass +
    A_IsOnEnzymes +
    treatment +
    (1 | eDWID),
  data = mod_df
)
t3 = car::Anova(m, type = 'III', contrasts = "contr.sum")
emm = emmeans(m, specs = ~ years_to_cfrd + a1c_ogtt, pbkrtest.limit = 1e5)
emm = data.frame(emm)
ggplot(
  emm,
  aes(x = years_to_cfrd, y = emmean, group = a1c_ogtt, color = a1c_ogtt)
) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.25) +
  geom_line() +
  xlab("Years From CFRD Dx") +
  ylab("Estimated Mean ppFEV1") +
  labs(color = "Diagnosis Method") +
  theme_bw()
# Model tables
kable(emm, digits = 3)
m |> tbl_regression() |> add_n(location = "level")
t3
```

@@

@@

### LOESS-smoothed plot

```{r}
#| warning: false
#| message: false
# Plot by diagnosis type
ggplot(
  mod_df,
  aes(
    y = A_FVC_pct_predicted,
    x = years_to_cfrd,
    color = a1c_ogtt,
    group = a1c_ogtt
  )
) +
  xlab("Years from CFRD") +
  geom_smooth() +
  theme_bw()
```

### Linear model results

#### Not adjusted for treatment status

Adjusted for age at diagnosis, sex, mutation class, and pancreatic enzyme use.

```{r}
# Fit the model
m <- lmer(
  A_FVC_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    age_dx +
    Gender +
    MutClass +
    A_IsOnEnzymes +
    (1 | eDWID),
  data = mod_df
)
t3 = car::Anova(m, type = 'III', contrasts = "contr.sum")
emm = emmeans(m, specs = ~ years_to_cfrd + a1c_ogtt, pbkrtest.limit = 1e5)
emm = data.frame(emm)
ggplot(
  emm,
  aes(x = years_to_cfrd, y = emmean, group = a1c_ogtt, color = a1c_ogtt)
) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.25) +
  geom_line() +
  xlab("Years From CFRD Dx") +
  ylab("Estimated Mean ppFVC") +
  labs(color = "Diagnosis Method") +
  theme_bw()
# Model tables
kable(emm, digits = 3)
m |> tbl_regression() |> add_n(location = "level")
t3
```

@@

#### Adjusted for treatment status

Adjusted for age at diagnosis, sex, mutation class, pancreatic enzyme use, and CFRD treatment.

```{r}
# Fit the model
m <- lmer(
  A_FVC_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    age_dx +
    Gender +
    MutClass +
    A_IsOnEnzymes +
    treatment +
    (1 | eDWID),
  data = mod_df
)
t3 = car::Anova(m, type = 'III', contrasts = "contr.sum")
emm = emmeans(m, specs = ~ years_to_cfrd + a1c_ogtt, pbkrtest.limit = 1e5)
emm = data.frame(emm)
ggplot(
  emm,
  aes(x = years_to_cfrd, y = emmean, group = a1c_ogtt, color = a1c_ogtt)
) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.25) +
  geom_line() +
  xlab("Years From CFRD Dx") +
  ylab("Estimated Mean ppFVC") +
  labs(color = "Diagnosis Method") +
  theme_bw()
# Model tables
kable(emm, digits = 3)
m |> tbl_regression() |> add_n(location = "level")
t3
```

@@

```{r}
# Write data for Christine to look at
mod_df |>
  select(
    eDWID,
    Race,
    Hispanicrace,
    MutClass,
    ReviewYear,
    Age_YrEnd,
    CFRD,
    first_cfrd,
    age_dx,
    A_weight,
    A_height,
    bmi_adult,
    bmi_perc,
    bmi_cat,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    a1c_ogtt,
    A_hgba1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    A_IsOnEnzymes,
    liver_disease,
    A_pseudomonasaeruginosa,
    A_fungalyeast1,
    A_mycocultureresults3,
    A_cultureresults3,
    treatment,
    cfrd_treatment_options2,
    cfrd_treatment_options3,
    cfrd_treatment_options4,
    years_with_data
  ) %>%
  write.csv(
    .,
    file = "./Data_Clean/fev1_model_data_2010.csv",
    row.names = FALSE,
    na = ""
  )
# Tables
mod_df |>
  filter(ReviewYear == first_cfrd) |>
  select(
    a1c_ogtt,
    age_dx,
    A_weight,
    A_height,
    bmi_adult,
    bmi_perc,
    bmi_cat,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    Race,
    Hispanicrace,
    MutClass,
    a1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    A_IsOnEnzymes,
    liver_disease,
    A_pseudomonasaeruginosa,
    A_fungalyeast1,
    A_mycocultureresults3,
    A_cultureresults3,
    treatment,
    cfrd_treatment_options2,
    cfrd_treatment_options3,
    cfrd_treatment_options4,
    years_with_data
  ) |>
  tbl_summary(
    by = a1c_ogtt,
    missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) |>
  add_overall() |>
  add_p(
    test = list(
      all_continuous() ~ "oneway.test",
      "years_with_data" ~ "kruskal.test"
    )
  ) |>
  modify_caption("**All Groups**")
mod_df |>
  filter(ReviewYear == first_cfrd) |>
  filter(a1c_ogtt %in% c("OGTT", "HbA1c Only")) |>
  mutate(a1c_ogtt = droplevels(a1c_ogtt)) |>
  select(
    a1c_ogtt,
    age_dx,
    A_weight,
    A_height,
    A_bmivalue,
    A_bmipercentile,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    Race,
    Hispanicrace,
    MutClass,
    a1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    A_IsOnEnzymes,
    liver_disease,
    A_pseudomonasaeruginosa,
    A_fungalyeast1,
    A_mycocultureresults3,
    A_cultureresults3,
    treatment,
    cfrd_treatment_options2,
    cfrd_treatment_options3,
    cfrd_treatment_options4,
    years_with_data
  ) |>
  tbl_summary(
    by = a1c_ogtt,
    missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) |>
  add_overall() |>
  add_p(test = list(all_continuous() ~ "oneway.test")) |>
  modify_caption("**OGTT vs. HbA1c Only**")
```

@@

### LOESS-smoothed plot

```{r}
#| warning: false
#| message: false
# Plot by diagnosis type
ggplot(
  mod_df,
  aes(
    y = A_FEV1_pct_predicted,
    x = years_to_cfrd,
    color = a1c_ogtt,
    group = a1c_ogtt
  )
) +
  xlab("Years from CFRD") +
  geom_smooth() +
  theme_bw()
```

### Linear model results

#### Not adjusted for treatment status

Adjusted for age at diagnosis, sex, mutation class, and pancreatic enzyme use.

```{r}
# Fit the model
m <- lmer(
  A_FEV1_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    age_dx +
    Gender +
    MutClass +
    A_IsOnEnzymes +
    (1 | eDWID),
  data = mod_df
)
t3 = car::Anova(m, type = 'III', contrasts = "contr.sum")
emm = emmeans(m, specs = ~ years_to_cfrd + a1c_ogtt, pbkrtest.limit = 1e5)
emm = data.frame(emm)
ggplot(
  emm,
  aes(x = years_to_cfrd, y = emmean, group = a1c_ogtt, color = a1c_ogtt)
) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.25) +
  geom_line() +
  xlab("Years From CFRD Dx") +
  ylab("Estimated Mean ppFEV1") +
  labs(color = "Diagnosis Method") +
  theme_bw()
# Model tables
kable(emm, digits = 3)
m |> tbl_regression() |> add_n(location = "level")
t3
```

@@

#### Adjusted for treatment status

Adjusted for age at diagnosis, sex, mutation class, pancreatic enzyme use, and CFRD treatment.

```{r}
# Fit the model
m <- lmer(
  A_FEV1_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    age_dx +
    Gender +
    MutClass +
    A_IsOnEnzymes +
    treatment +
    (1 | eDWID),
  data = mod_df
)
t3 = car::Anova(m, type = 'III', contrasts = "contr.sum")
emm = emmeans(m, specs = ~ years_to_cfrd + a1c_ogtt, pbkrtest.limit = 1e5)
emm = data.frame(emm)
ggplot(
  emm,
  aes(x = years_to_cfrd, y = emmean, group = a1c_ogtt, color = a1c_ogtt)
) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.25) +
  geom_line() +
  xlab("Years From CFRD Dx") +
  ylab("Estimated Mean ppFEV1") +
  labs(color = "Diagnosis Method") +
  theme_bw()
# Model tables
kable(emm, digits = 3)
m |> tbl_regression() |> add_n(location = "level")
t3
```

>>>>

@@

```{r}
# Create a model dataframe
mod_df = inner_join(
  annual,
  cfrd_only |> filter(dx_year >= 2004),
  by = join_by(eDWID)
) |>
  drop_na(a1c_ogtt) |>
  mutate(years_to_cfrd = ReviewYear - first_cfrd) |>
  filter(years_to_cfrd <= 5 & years_to_cfrd >= -5) |>
  mutate(years_to_cfrd = factor(years_to_cfrd))
var_label(mod_df$years_to_cfrd) = "Years From CFRD Dx"
# Write data for Christine to look at
mod_df |>
  select(
    eDWID,
    Race,
    Hispanicrace,
    MutClass,
    ReviewYear,
    Age_YrEnd,
    CFRD,
    first_cfrd,
    age_dx,
    A_weight,
    A_height,
    bmi_adult,
    bmi_perc,
    bmi_cat,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    a1c_ogtt,
    A_hgba1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    A_IsOnEnzymes,
    liver_disease,
    A_pseudomonasaeruginosa,
    A_fungalyeast1,
    A_mycocultureresults3,
    A_cultureresults3,
    treatment,
    cfrd_treatment_options2,
    cfrd_treatment_options3,
    cfrd_treatment_options4,
    NumFEV1,
    NumFVC
  ) %>%
  write.csv(
    .,
    file = "./Data_Clean/fev1_model_data_2004.csv",
    row.names = FALSE,
    na = ""
  )
# Tables
mod_df |>
  filter(ReviewYear == first_cfrd) |>
  select(
    a1c_ogtt,
    age_dx,
    A_weight,
    A_height,
    bmi_adult,
    bmi_perc,
    bmi_cat,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    Race,
    Hispanicrace,
    MutClass,
    a1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    A_IsOnEnzymes,
    liver_disease,
    A_pseudomonasaeruginosa,
    A_fungalyeast1,
    A_mycocultureresults3,
    A_cultureresults3,
    treatment,
    cfrd_treatment_options2,
    cfrd_treatment_options3,
    cfrd_treatment_options4,
    NumFEV1,
    NumFVC
  ) |>
  tbl_summary(
    by = a1c_ogtt,
    missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) |>
  add_overall() |>
  add_p() |>
  modify_caption("**All Groups**")
mod_df |>
  filter(ReviewYear == first_cfrd, a1c_ogtt %in% c("OGTT", "HbA1c Only")) |>
  mutate(a1c_ogtt = droplevels(a1c_ogtt)) |>
  select(
    a1c_ogtt,
    age_dx,
    A_weight,
    A_height,
    A_bmivalue,
    A_bmipercentile,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    Race,
    Hispanicrace,
    MutClass,
    a1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    A_IsOnEnzymes,
    liver_disease,
    A_pseudomonasaeruginosa,
    A_fungalyeast1,
    A_mycocultureresults3,
    A_cultureresults3,
    treatment,
    cfrd_treatment_options2,
    cfrd_treatment_options3,
    cfrd_treatment_options4,
    NumFEV1,
    NumFVC
  ) |>
  tbl_summary(
    by = a1c_ogtt,
    missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) |>
  add_overall() |>
  add_p(test = list(all_continuous() ~ "oneway.test")) |>
  modify_caption("**OGTT vs. HbA1c Only**")
```

@@

### Not adjusted for treatment status

Adjusted for age at diagnosis, number of FEV1 measures, sex, mutation class, and pancreatic enzyme use.

```{r}
# Fit the model
m <- lmer(
  A_FEV1_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    age_dx +
    NumFEV1 +
    Gender +
    MutClass +
    A_IsOnEnzymes +
    (1 | eDWID),
  data = mod_df
)
t3 = car::Anova(m, type = 'III', contrasts = "contr.sum")
emm = emmeans(m, specs = ~ years_to_cfrd + a1c_ogtt, pbkrtest.limit = 1e5)
ggplot(
  data.frame(emm),
  aes(x = years_to_cfrd, y = emmean, group = a1c_ogtt, color = a1c_ogtt)
) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.25) +
  geom_line() +
  xlab("Years From CFRD Dx") +
  ylab("Estimated Mean ppFEV1") +
  labs(color = "Diagnosis Method") +
  theme_bw()
# Model tables
m |> tbl_regression() |> add_n(location = "level")
```

@@

```{r}
# Contrasts - probably a more elegant way to do this but oh well
neither_m_5 = rep(0, times = 33)
neither_m_5[1] = 1
neither_m_1 = rep(0, times = 33)
neither_m_1[5] = 1
neither_0 = rep(0, times = 33)
neither_0[6] = 1
neither_1 = rep(0, times = 33)
neither_1[7] = 1
neither_5 = rep(0, times = 33)
neither_5[11] = 1
hba1c_m_5 = rep(0, times = 33)
hba1c_m_5[12] = 1
hba1c_m_1 = rep(0, times = 33)
hba1c_m_1[16] = 1
hba1c_0 = rep(0, times = 33)
hba1c_0[17] = 1
hba1c_1 = rep(0, times = 33)
hba1c_1[18] = 1
hba1c_5 = rep(0, times = 33)
hba1c_5[22] = 1
ogtt_m_5 = rep(0, times = 33)
ogtt_m_5[23] = 1
ogtt_m_1 = rep(0, times = 33)
ogtt_m_1[27] = 1
ogtt_0 = rep(0, times = 33)
ogtt_0[28] = 1
ogtt_1 = rep(0, times = 33)
ogtt_1[29] = 1
ogtt_5 = rep(0, times = 33)
ogtt_5[33] = 1
# Test contrasts
F <- contrast(
  emm,
  list(
    "Neither: -5 to 0 years" = neither_0 - neither_m_5,
    "Neither: 0 to 5 years" = neither_5 - neither_0,
    "Neither: -1 to 0 years" = neither_0 - neither_m_1,
    "Neither: 0 to 1 years" = neither_1 - neither_0,
    "HbA1c Only: -5 to 0 years" = hba1c_0 - hba1c_m_5,
    "HbA1c Only: 0 to 5 years" = hba1c_5 - hba1c_0,
    "HbA1c Only: -1 to 0 years" = hba1c_0 - hba1c_m_1,
    "HbA1c Only: 0 to 1 years" = hba1c_1 - hba1c_0,
    "OGTT: -5 to 0 years" = ogtt_0 - ogtt_m_5,
    "OGTT: 0 to 5 years" = ogtt_5 - ogtt_0,
    "OGTT: -1 to 0 years" = ogtt_0 - ogtt_m_1,
    "OGTT: 0 to 1 years" = ogtt_1 - ogtt_0
  )
)
kable(F, digits = 3)
kable(data.frame(emm), digits = 3)
```

The table above indicates the change in average lung function from a given timepoint to another within each group. For example, a value of -13.9 for "Neither: -5 to 0 years" indicates that among people without HbA1c or OGTT, lung function declined by 13 percentage points on average from 5 years prior to diagnosis to time of diagnosis.

### Adjusted for treatment status

Adjusted for age at diagnosis, number of FEV1 measures, sex, mutation class, pancreatic enzyme use, and CFRD treatment.

```{r}
# Fit the model
m <- lmer(
  A_FEV1_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    age_dx +
    NumFEV1 +
    Gender +
    MutClass +
    A_IsOnEnzymes +
    treatment +
    (1 | eDWID),
  data = mod_df
)
t3 = car::Anova(m, type = 'III', contrasts = "contr.sum")
emm = emmeans(m, specs = ~ years_to_cfrd + a1c_ogtt, pbkrtest.limit = 1e5)
ggplot(
  data.frame(emm),
  aes(x = years_to_cfrd, y = emmean, group = a1c_ogtt, color = a1c_ogtt)
) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.25) +
  geom_line() +
  xlab("Years From CFRD Dx") +
  ylab("Estimated Mean ppFEV1") +
  labs(color = "Diagnosis Method") +
  theme_bw()
# Model table
m |> tbl_regression() |> add_n(location = "level")
```

@@

```{r}
# Test contrasts
F <- contrast(
  emm,
  list(
    "Neither: -5 to 0 years" = neither_0 - neither_m_5,
    "Neither: 0 to 5 years" = neither_5 - neither_0,
    "Neither: -1 to 0 years" = neither_0 - neither_m_1,
    "Neither: 0 to 1 years" = neither_1 - neither_0,
    "HbA1c Only: -5 to 0 years" = hba1c_0 - hba1c_m_5,
    "HbA1c Only: 0 to 5 years" = hba1c_5 - hba1c_0,
    "HbA1c Only: -1 to 0 years" = hba1c_0 - hba1c_m_1,
    "HbA1c Only: 0 to 1 years" = hba1c_1 - hba1c_0,
    "OGTT: -5 to 0 years" = ogtt_0 - ogtt_m_5,
    "OGTT: 0 to 5 years" = ogtt_5 - ogtt_0,
    "OGTT: -1 to 0 years" = ogtt_0 - ogtt_m_1,
    "OGTT: 0 to 1 years" = ogtt_1 - ogtt_0
  )
)
kable(F, digits = 3)
kable(data.frame(emm), digits = 3)
```

The table above indicates the change in average lung function from a given timepoint to another within each group. For example, a value of -13.9 for "Neither: -5 to 0 years" indicates that among people without HbA1c or OGTT, lung function declined by 13 percentage points on average from 5 years prior to diagnosis to time of diagnosis.

@@

### Not adjusted for treatment status

Adjusted for age at diagnosis, number of FVC measures, sex, mutation class, and pancreatic enzyme use.

```{r}
# Fit the model
m <- lmer(
  A_FVC_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    age_dx +
    NumFVC +
    Gender +
    MutClass +
    A_IsOnEnzymes +
    (1 | eDWID),
  data = mod_df
)
t3 = car::Anova(m, type = 'III', contrasts = "contr.sum")
emm = emmeans(m, specs = ~ years_to_cfrd + a1c_ogtt, pbkrtest.limit = 1e5)
ggplot(
  data.frame(emm),
  aes(x = years_to_cfrd, y = emmean, group = a1c_ogtt, color = a1c_ogtt)
) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.25) +
  geom_line() +
  xlab("Years From CFRD Dx") +
  ylab("Estimated Mean ppFVC") +
  labs(color = "Diagnosis Method") +
  theme_bw()
# Model tables
m |> tbl_regression() |> add_n(location = "level")
```

@@

```{r}
# Test contrasts
F <- contrast(
  emm,
  list(
    "Neither: -5 to 0 years" = neither_0 - neither_m_5,
    "Neither: 0 to 5 years" = neither_5 - neither_0,
    "Neither: -1 to 0 years" = neither_0 - neither_m_1,
    "Neither: 0 to 1 years" = neither_1 - neither_0,
    "HbA1c Only: -5 to 0 years" = hba1c_0 - hba1c_m_5,
    "HbA1c Only: 0 to 5 years" = hba1c_5 - hba1c_0,
    "HbA1c Only: -1 to 0 years" = hba1c_0 - hba1c_m_1,
    "HbA1c Only: 0 to 1 years" = hba1c_1 - hba1c_0,
    "OGTT: -5 to 0 years" = ogtt_0 - ogtt_m_5,
    "OGTT: 0 to 5 years" = ogtt_5 - ogtt_0,
    "OGTT: -1 to 0 years" = ogtt_0 - ogtt_m_1,
    "OGTT: 0 to 1 years" = ogtt_1 - ogtt_0
  )
)
kable(F, digits = 3)
kable(data.frame(emm), digits = 3)
```

The table above indicates the change in average lung function from a given timepoint to another within each group. For example, a value of -13.9 for "Neither: -5 to 0 years" indicates that among people without HbA1c or OGTT, lung function declined by 13 percentage points on average from 5 years prior to diagnosis to time of diagnosis.

### Adjusted for treatment status

Adjusted for age at diagnosis, number of FVC measures, sex, mutation class, pancreatic enzyme use, and CFRD treatment.

```{r}
# Fit the model
m <- lmer(
  A_FVC_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    age_dx +
    NumFVC +
    Gender +
    MutClass +
    A_IsOnEnzymes +
    treatment +
    (1 | eDWID),
  data = mod_df
)
t3 = car::Anova(m, type = 'III', contrasts = "contr.sum")
emm = emmeans(m, specs = ~ years_to_cfrd + a1c_ogtt, pbkrtest.limit = 1e5)
ggplot(
  data.frame(emm),
  aes(x = years_to_cfrd, y = emmean, group = a1c_ogtt, color = a1c_ogtt)
) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.25) +
  geom_line() +
  xlab("Years From CFRD Dx") +
  ylab("Estimated Mean ppFVC") +
  labs(color = "Diagnosis Method") +
  theme_bw()
# Model tables
m |> tbl_regression() |> add_n(location = "level")
```

@@

```{r}
# Write data for Christine to look at
mod_df |>
  select(
    eDWID,
    Race,
    Hispanicrace,
    MutClass,
    ReviewYear,
    Age_YrEnd,
    CFRD,
    first_cfrd,
    age_dx,
    A_weight,
    A_height,
    bmi_adult,
    bmi_perc,
    bmi_cat,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    a1c_ogtt,
    A_hgba1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    A_IsOnEnzymes,
    liver_disease,
    A_pseudomonasaeruginosa,
    A_fungalyeast1,
    A_mycocultureresults3,
    A_cultureresults3,
    treatment,
    cfrd_treatment_options2,
    cfrd_treatment_options3,
    cfrd_treatment_options4,
    NumFEV1,
    NumFVC
  ) %>%
  write.csv(
    .,
    file = "./Data_Clean/fev1_model_data_2010.csv",
    row.names = FALSE,
    na = ""
  )
# Tables
mod_df |>
  filter(ReviewYear == first_cfrd) |>
  select(
    a1c_ogtt,
    age_dx,
    A_weight,
    A_height,
    bmi_adult,
    bmi_perc,
    bmi_cat,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    Race,
    Hispanicrace,
    MutClass,
    a1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    A_IsOnEnzymes,
    liver_disease,
    A_pseudomonasaeruginosa,
    A_fungalyeast1,
    A_mycocultureresults3,
    A_cultureresults3,
    treatment,
    cfrd_treatment_options2,
    cfrd_treatment_options3,
    cfrd_treatment_options4,
    NumFEV1,
    NumFVC
  ) |>
  tbl_summary(
    by = a1c_ogtt,
    missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) |>
  add_overall() |>
  add_p() |>
  modify_caption("**All Groups**")
mod_df |>
  filter(ReviewYear == first_cfrd) |>
  filter(a1c_ogtt %in% c("OGTT", "HbA1c Only")) |>
  mutate(a1c_ogtt = droplevels(a1c_ogtt)) |>
  select(
    a1c_ogtt,
    age_dx,
    A_weight,
    A_height,
    A_bmivalue,
    A_bmipercentile,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    Race,
    Hispanicrace,
    MutClass,
    a1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    A_IsOnEnzymes,
    liver_disease,
    A_pseudomonasaeruginosa,
    A_fungalyeast1,
    A_mycocultureresults3,
    A_cultureresults3,
    treatment,
    cfrd_treatment_options2,
    cfrd_treatment_options3,
    cfrd_treatment_options4,
    NumFEV1,
    NumFVC
  ) |>
  tbl_summary(
    by = a1c_ogtt,
    missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) |>
  add_overall() |>
  add_p(test = list(all_continuous() ~ "oneway.test")) |>
  modify_caption("**OGTT vs. HbA1c Only**")
```

@@

### Not adjusted for treatment status

Adjusted for age at diagnosis, number of FEV1 measures, sex, mutation class, and pancreatic enzyme use.

```{r}
# Fit the model
m <- lmer(
  A_FEV1_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    age_dx +
    NumFEV1 +
    Gender +
    MutClass +
    A_IsOnEnzymes +
    (1 | eDWID),
  data = mod_df
)
t3 = car::Anova(m, type = 'III', contrasts = "contr.sum")
emm = emmeans(m, specs = ~ years_to_cfrd + a1c_ogtt, pbkrtest.limit = 1e5)
ggplot(
  data.frame(emm),
  aes(x = years_to_cfrd, y = emmean, group = a1c_ogtt, color = a1c_ogtt)
) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.25) +
  geom_line() +
  xlab("Years From CFRD Dx") +
  ylab("Estimated Mean ppFEV1") +
  labs(color = "Diagnosis Method") +
  theme_bw()
# Model table
m |> tbl_regression() |> add_n(location = "level")
```

@@

```{r}
# Test contrasts
F <- contrast(
  emm,
  list(
    "Neither: -5 to 0 years" = neither_0 - neither_m_5,
    "Neither: 0 to 5 years" = neither_5 - neither_0,
    "Neither: -1 to 0 years" = neither_0 - neither_m_1,
    "Neither: 0 to 1 years" = neither_1 - neither_0,
    "HbA1c Only: -5 to 0 years" = hba1c_0 - hba1c_m_5,
    "HbA1c Only: 0 to 5 years" = hba1c_5 - hba1c_0,
    "HbA1c Only: -1 to 0 years" = hba1c_0 - hba1c_m_1,
    "HbA1c Only: 0 to 1 years" = hba1c_1 - hba1c_0,
    "OGTT: -5 to 0 years" = ogtt_0 - ogtt_m_5,
    "OGTT: 0 to 5 years" = ogtt_5 - ogtt_0,
    "OGTT: -1 to 0 years" = ogtt_0 - ogtt_m_1,
    "OGTT: 0 to 1 years" = ogtt_1 - ogtt_0
  )
)
kable(F, digits = 3)
kable(data.frame(emm), digits = 3)
```

The table above indicates the change in average lung function from a given timepoint to another within each group. For example, a value of -13.9 for "Neither: -5 to 0 years" indicates that among people without HbA1c or OGTT, lung function declined by 13 percentage points on average from 5 years prior to diagnosis to time of diagnosis.

### Adjusted for treatment status

Adjusted for age at diagnosis, number of FEV1 measures, sex, mutation class, pancreatic enzyme use, and CFRD treatment.
